---
title: "BGGN 213 Lecture 8"
author: "Eduardo Ramirez"
date: "4/27/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clustering example
# cbind converts vectors into matrixes, binding with rev
Make up some data for testing "Kmeans()" function

```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

Use the kmeans() function setting k to 2 and nstart=20
```{r}
km <- kmeans(x, centers = 2, nstart=20)
km
```
```{r}
km$size
```
```{r}
km$cluster
```
```{r}
km$centers
```

Inspect/print the results

Q. How many points are in each cluster?
Q. What ‘component’ of your result object details
      - cluster size?
      - cluster assignment/membership?
      - cluster center?
Plot x colored by the kmeans cluster assignment and
      add cluster centers as blue points
```{r}
plot(x, col=km$cluster, pch=16)
points(km$centers, col="blue", pch=15)




```
```{r}
 First we need to calculate point (dis)similarity
#   as the Euclidean distance between observations
dist_matrix <- dist(x)
# The hclust() function returns a hierarchical
#  clustering model
hc <- hclust(d = dist_matrix)
# the print method is not so useful here
hc
```
      # First we need to calculate point (dis)similarity
#   as the Euclidean distance between observations
dist_matrix <- dist(x)
# The hclust() function returns a hierarchical
#  clustering model
hc <- hclust(d = dist_matrix)
# the print method is not so useful here
hc
```{r}
dist_matrix <- dist(x)
```
      
      hc - hclust(d = dist_matrix)
```{r}
 hc <- hclust(d = dist_matrix)
```
```{r}
hc
```
```{r}
#Convert to matrix to see structure of matrix and dimenstions
dim(as.matrix(dist_matrix))
```
```{r}
dist_matrix <- dist(x)
hc <- hclust(d = dist_matrix)
hc
```
```{r}
class(hc)
plot(hc)
```
```{r}
plot(hc)
grps <- abline(h=6, col="red")
table(grps)
```
```{r}
plot(hc)
abline(h=6, col="red")
cutree(hc, k=2)
```
```{r}
plot(x, col=grps)
```
Try different cutting
```{r}
plot(x, col=cutree(hc, k=4))
```
```{r}
hc.complete <- hclust(d, method="complete")
hc.average  <- hclust(d, method="average")
hc.single   <- hclust(d, method="single")
```
## Another hclust example
```{r}
x <- rbind(
  matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2),   # c1
  matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
  matrix(c(rnorm(50, mean = 1, sd = 0.3),           # c3
           rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")

```
Plotting x
```{r}
plot(x)
```
```{r}
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col)
```
# hclust() to cut data
```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```
Cut into 2 groups
```{r}
grps <- cutree(hc, k=3)
table(grps)
```
Plot data colored by clusters
```{r}
plot(x, col = grps)
```
# Importing data
```{r}
mydata <- read.csv("https://tinyurl.com/expression-CSV", row.names = 1)
head(mydata)
```
Invert data (vertical vs horizontal)
```{r}
View(t(mydata))
```

```{r}
 ## lets do PCA
pca <- prcomp(t(mydata), scale=TRUE)
## See what is returned by the prcomp() function
   attributes(pca)
```
# Not useful
```{r}
pca
```

```{r}
 ## lets do PCA
pca <- prcomp(t(mydata), scale=TRUE)
## A basic PC1 vs PC2 2-D plot
plot(pca$x[,1], pca$x[,2])
## Variance captured per PC
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```
Plotting (Scree plot)
```{r}
 pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
barplot(pca.var.per, main="Scree Plot",
        xlab="Principal Component", ylab="Percent Variation")
```
Color up PCA plot
```{r}
as.factor(colnames(mydata))
```

```{r}
as.factor(substr(colnames(mydata), 1, 2))
mycols <- as.factor(substr(colnames(mydata), 1, 2))
```
```{r}

```
#pca.var.per is important for inputting real numbers from the data
```{r}
 ## A vector of colors for wt and ko samples
colvec <- as.factor( substr( colnames(mydata), 1, 2) )
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))
```

```{r}
paste0("PC1 (",pca.var.per[1],"%)")
plot(pca$x[,1], pca$x[,2], col=mycols, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))
   
```
```{r}
paste0("PC1 (",pca.var.per[1],"%)")
plot(pca$x[,1], pca$x[,2], col=mycols, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))
text(pca$x[,1], pca$x[,2], labels=colnames(mydata), pos=1)
```

```{r}
x <- read.csv("UK_foods.csv")
dim(x)
head(x)
```

