---
title: "BGGN 213 Lecture"
author: "Eduardo Ramirez"
date: "5/2/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
url <- "https://bioboot.github.io/bggn213_S18/class-material/WisconsinCancer.csv"
wisc.df <- read.csv(url)
# Complete the following code to input the data and store as wisc.df


head(wisc.df)

```

How many diagnosis are cancer vs noncancer
```{r}
table(wisc.df$diagnosis)
```

```{r}
# Convert the features of the data: wisc.data
#wisc.data <- as.matrix( ___ )
#x <- c("barry, "chris", "mary", "pete)
#x[-c(1,4)]
###Removing the problem of 'X' column in position 33!
wisc.data <- as.matrix(wisc.df[,-c(1:2,33)])
rownames(wisc.data) <- wisc.df$id
head(wisc.data)

x <- c("barry", "chris", "mary", "pete")
x[-c(1,4)]
```
Create diagnosis vector
```{r}
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
sum(diagnosis)
```
###Questions
Explore the data you created before (wisc.data and diagnosis) to answer the following questions:

Q1. How many observations are in this dataset? 
#Use dimmensions (rows), (nrow) -> 569 observations
Q2. How many variables/features in the data are suffixed with _mean?
#31 columns
Q3. How many of the observations have a malignant diagnosis?
#Sum of diagnosis -> 212
The functions dim(), length(), grep() and sum() may be useful for answering the first 3 questions above.
#suffixed = at the end
```{r}
nrow(wisc.data)
ncol(wisc.data)
table(wisc.df$diagnosis)
```
```{r}
colnames(wisc.data)
```
#grep: most important = pattern/x
```{r}
#grep("_mean", colnames(wisc.data), value=TRUE, invert=TRUE)
length(grep("_mean", colnames(wisc.data)))
```
The functions dim(), length(), grep() and sum() may be useful for answering the first 3 questions above.
#Principal Component Analysis
Check the mean and standard deviation of the features (i.e. columns) of the wisc.data to determine if the data should be scaled. Use the colMeans() and apply() functions like you’ve done before.
```{r}
colMeans(wisc.data)
plot(colMeans(wisc.data), type = "o")
apply(wisc.data,2,sd)
```

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale. = TRUE)
#Now Column 33 is removed from our wisc.data
summary(wisc.pr)
```
##PCA result plots
```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = diagnosis+3)
wisc.pr$sdev^2
```
##Scree-plot: variance explained
```{r}
# Variance explained by each principal component: pve
pve <- wisc.pr$sdev^2 / sum(wisc.pr$sdev^2)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```
Barplot of data
```{r}
barplot(pve, names.arg = paste("PC", 1:length(pve)), las=2, axes = FALSE, ylab = "Precent of Variance Explained")
axis(2, at=pve, labels = round(pve,2)*100)
```

#Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

#Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

#Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

#Hierarchical clustering
Scale date using scale() function
```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)
```
Calculate the (Euclidean) distances between all pairs of observations in the new scaled dataset and assign the result to data.dist.
```{r}
data.dist <- dist(data.scaled)
```
Create a hierarchical clustering model using complete linkage. Manually specify the method argument to hclust() and assign the results to wisc.hclust.
```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```
Results of hierarchical clustering
```{r}
plot(wisc.hclust)
abline(h=20, col="red", lwd=3)
```
#Cut Tree to remove clusters
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```
#How well do groups match diagnosis
#We can use the table() function to compare the cluster membership to the actual diagnoses.
```{r}
table(diagnosis)
table(, diagnosis)
```
##K-means clustering and comparing results
#Compare to diagnosis
```{r}
wisc.km <- kmeans(data.scaled, centers= 2, nstart= 20)
table(wisc.km$cluster)
table(wisc.km$cluster, diagnosis)
```
##Clustering on PCA results
```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:3]), method="ward.D2")
plot(wisc.pr.hclust)
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=4)
table(wisc.pr.hclust.clusters, diagnosis)
plot(wisc.pr$x[,1:2], col=wisc.pr.hclust.clusters)
```
```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```
###Installing rgl
```{r}
#install.packages("rgl")
library(rgl)

```

###Bonus Section
##Predicting Malignancy of New Samples
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
```

```{r}
plot(wisc.pr$x[,1:2], col = wisc.pr.hclust.clusters)
points(npc[,1], npc[,2], col=c("purple", "blue", pch=16, cex=3)
```

